Anthropomorphism in Human-Robot Interaction: Nature, Dynamics and Underlying Cognitive Mechanisms

### 2014-11-25 paper.tex
Section 3: {stage->layer}

Finally, we synthetize and integrate in an explanatory model the \emph{cognitive
mechanisms} that underlie anthropomorphism. Several studies previously showed
that we build instinctive bonds with robots as soon as we attribute animacy, and
we also know from social psychology some of the mechanisms that lead to
\emph{familiarization}. We attempt here to build a practical model of the
cognitive stages we encounter in human-robot interaction.

@@

This abstract model may contribute to a better understanding of the
\emph{phenomenology} of anthropomorphism, but it does not help to \emph{explain}
the phenomenon. This section proposes to look at the dynamics of
anthropomorphism in terms of \emph{cognitive correlates}: how human cognition
can explain the dynamic nature of our interactions with robots, and what are the
main cognitive stages upon which researchers can build to design better, deeper
human-robot interactions.

@@

\subsection{A Model of the Cognitive Stages of Human-Robot Interaction}
\label{sec:cognitive-model}

@@

\draw (-0.3,2) node[anchor=east] {Stage 1}
      (-0.3,4) node[anchor=east] {Stage 2}
      (-0.3,6) node[anchor=east] {Stage 3};

@@

\end{tikzpicture}
%}
\caption{\textbf{The three cognitive stages of anthropomorphism}: Stage 1 is the instinctive,
sub-cognitive identification of living peers. {\it Empathy} is characteristic
of this stage. After longer observation or short, non-contextualized interaction
(typically, a lab environment), the user reaches Stage 2: the user projects a
mental model he/she is already familiar with onto the robot. After longer {\it
contextualized} interaction (typically, at home), the user enters Stage 3 of
anthropomorphism: the user composes a custom mental model of the robot,
based on experience. This leads to adapted interaction modalities, that may
still be anthropomorphic, or not. Note that the different stages coexist: Stage
1 and 2 still affect the human perception of the robot in a Stage 3 interaction.}
\label{fig:cognitivemodel}
\end{figure}

We propose three different cognitive stages (Figure~\ref{fig:cognitivemodel}),
which are not mutually exclusive but overlap over time. Note that these three
cognitive stages are related but do not exactly match the \emph{Initialization},
\emph{Familiarization} and \emph{Stabilization} phases introduced in our model
of the dynamics of anthropomorphism. We discuss this relation hereafter.

\subsubsection{Cognitive Processes and Stages\\}

@@

The \emph{Stage 1} is the instinctive, pre-cognitive identification of living
peers. Studies conducted by~\citet{rosenthal-vonderputten_experimental_2013},
who investigated the neural correlates of emotional reactions of humans towards
a robot, supports the idea that humans tend to anthropomorphize robots
intuitively in this pre-cognitive way. {\it Empathy} is characteristic of this
stage~\citep{rosenthalvonderPutten2013neural}.  It is also at this stage that
automatic activation of the human's mirror neurons system  in response to
viewing the robots action and human automatic emotional responses are
contributing at a lower level to the intuitive anthropomorphization.

The intuitive anthropomorphization in the cognitive Stage 1 might be
characteristic of what \cite{takayama_perspectives_2012} calls the
\textit{in-the-moment} perspective on agency with a robot compared to the
\textit{reflective} perspective (see paragraph ``artifact-centric perspective''
in section \ref{sec:anthropomorphism}).  For instance, we reflectively might
\textit{not} perceive agency in a social robot, but it can feel quite
differently in the moment of interaction. \cite{takayama_perspectives_2012}
argues that neglecting to separate reflective perspectives from in-the-moment
perspectives of agency is one of the major sources of confusion when people talk
and write about anthropomorphism. There seems to be a disconnection between what
people consciously perceive and how they respond to stimuli that they may not
consciously perceive. This means that in the initial phase of interacting with a
novel device \textit{(initialization)} (see Figure \ref{fig:dynamics}) people
might not respond consciously but rather mindlessly \citep{nass_machines_2000},
and in turn anthropomorphize more. They start to respond in a more reflective
manner only during the \textit{familiarization} (in turn, the tendency to
anthropomorphize might decrease). This can be illustrated by the fact that
people tend to deny interacting with computational systems as if they were
people and yet they respond to computers in many ways that are remarkably
similar to how they respond to people \citep{reeves_media_1996}.
\cite{takayama_perspectives_2012} also applies a cognitive viewpoint on the two
different perspectives of in-the-moment \vs reflective to illustrate
differences. She states that in-the-moment perceptions of agency are largely
shaped by bottom-up perceptual processes, evoking very immediate responses (\eg
to the cues emitted by the artifact's design (``artifact-centric
perspective''). In contrast, according to Takayama, reflective perceptions are
more often shaped by top-down processes because of the nature of reflective
thought.

After a longer observation period (typically including complete action sequences
of the robot) or short interaction (touching, short talk like greetings), we
suggest the human reaches the cognitive \emph{Stage 2}: in this phase, the human
starts building a behavioral and cognitive model of the robot that would support
both the observed and imagined capabilities of the robot.  The \emph{familiarity
thesis}~\citep{hegel_understanding_2008} supports the idea that the human first
projects onto the robot mental models of similar agents he/she is already
familiar with (ranging from animals to human adults, to pets and children). We
hypothesize that the nature of the projected mental model, as well as how deep
the human engages in this projection, might be driven by the same parameters as
we mentioned for the \emph{initial potential to anthropomorphize}. We also believe
that, at this stage, the human also changes his attitude towards the robot, by
focusing his attention to social cue more than to low-level behaviors, which
may in turn reinforce the resonance mechanisms.

The cognitive \emph{Stage 3} is reached after a \emph{contextualized} interaction.
A \emph{contextualized} interaction is \emph{explicitly purposeful} (the purpose
of the interaction, be it purely entertainment, is explicit and conscious to the
human), and takes place in an environment that fosters a stronger cognitive (and
possibly affective/social) commitment from the human in the interaction
(typically, at home). During this interaction, the human iteratively restates
and reshapes his/her behavioral and mental model of the robot (\emph{How does
the robot react to such and such situation/input?  What does the robot know
about me? About itself? About our environment? What can the robot learn?}, etc.).

@@

\draw (-0.3,1.5) node[anchor=east] {Stage 1}
      (-0.3,3) node[anchor=east] {Stage 2}
      (-0.3,4.5) node[anchor=east] {Stage 3};

@@

As mentioned before, the three cognitive stages proposed here do not directly
match with the three phases in the model of the dynamics of anthropomorphism,
and Figure~\ref{fig:mix} attempts to depict their interplay.
The cognitive Stages 1 and 2, in particular, are both active in the
\emph{initialization} phase of the anthropomorphism model. Sub-cognitive
anthropomorphism typically \emph{initiates} the novelty effect by rapidly
engaging the human in the interaction through an initial projected agency,
whereas cognitive Stage 2 (projection of familiar mental models) supports the
novelty effect by inducing beliefs that the robot is set up with possibly
complex cognitive abilities.

The cognitive Stage 3 also overlaps with the \emph{Familiarization} phase: as
the human gets used to the robot, he/she restates and adapts its
cognitive model of the robot by iteratively reshaping pre-existent, familiar
models until it provides a satisfying support to explain and justify the
observed robot behavior.

A \emph{stable level of anthropomorphism} is reached when the adaptation process
depicted in the cognitive Stage 3 reached a stable state, \ie the user's
experience with the robot is correctly supported by the cognitive model he/she
has built.

>>>>

Finally, we synthetize and integrate in an explanatory model the \emph{cognitive
mechanisms} that underlie anthropomorphism. Several studies previously showed
that we build instinctive bonds with robots as soon as we attribute animacy, and
we also know from social psychology some of the mechanisms that lead to
\emph{familiarization}. We attempt here to build a practical model of the
cognitive layers we encounter in human-robot interaction.

@@

This abstract model may contribute to a better understanding of the
\emph{phenomenology} of anthropomorphism, but it does not help to \emph{explain}
the phenomenon. This section proposes to look at the dynamics of
anthropomorphism in terms of \emph{cognitive correlates}: how human cognition
can explain the dynamic nature of our interactions with robots, and what are the
main cognitive layers upon which researchers can build to design better, deeper
human-robot interactions.

@@

\subsection{A Model of the Cognitive Layers of Human-Robot Interaction}
\label{sec:cognitive-model}

@@

\draw (-0.3,2) node[anchor=east] {Layer 1}
      (-0.3,4) node[anchor=east] {Layer 2}
      (-0.3,6) node[anchor=east] {Layer 3};

@@

\end{tikzpicture}
%}
\caption{\textbf{The three cognitive layers of anthropomorphism}: Layer 1 is the instinctive,
sub-cognitive identification of living peers. {\it Empathy} is characteristic
of this stage. After longer observation or short, non-contextualized interaction
(typically, a lab environment), the user enters Layer 2: the user projects a
mental model he/she is already familiar with onto the robot. After longer {\it
contextualized} interaction (typically, at home), the user enters Layer 3 of
anthropomorphism: the user composes a custom mental model of the robot,
based on experience. This leads to adapted interaction modalities, that may
still be anthropomorphic, or not. Note that the different layers coexist: Layer
1 and 2 still affect the human perception of the robot in a Layer 3 interaction.}
\label{fig:cognitivemodel}
\end{figure}

We propose three different cognitive layers (Figure~\ref{fig:cognitivemodel}),
which are not mutually exclusive but overlap over time. Note that these three
cognitive stages are related but do not exactly match the \emph{Initialization},
\emph{Familiarization} and \emph{Stabilization} phases introduced in our model
of the dynamics of anthropomorphism. We discuss this relation hereafter.

\subsubsection{Cognitive Processes and Layers\\}

@@

The \emph{Layer 1} is the instinctive, pre-cognitive identification of living
peers. Studies conducted by~\citet{rosenthal-vonderputten_experimental_2013},
who investigated the neural correlates of emotional reactions of humans towards
a robot, supports the idea that humans tend to anthropomorphize robots
intuitively in this pre-cognitive way. {\it Empathy} is characteristic of this
layer~\citep{rosenthalvonderPutten2013neural}.  It is also at this stage that
automatic activation of the human's mirror neurons system  in response to
viewing the robots action and human automatic emotional responses are
contributing at a lower level to the intuitive anthropomorphization.

The intuitive anthropomorphization in the cognitive Layer 1 might be
characteristic of what \cite{takayama_perspectives_2012} calls the
\textit{in-the-moment} perspective on agency with a robot compared to the
\textit{reflective} perspective (see paragraph ``artifact-centric perspective''
in section \ref{sec:anthropomorphism}).  For instance, we reflectively might
\textit{not} perceive agency in a social robot, but it can feel quite
differently in the moment of interaction. \cite{takayama_perspectives_2012}
argues that neglecting to separate reflective perspectives from in-the-moment
perspectives of agency is one of the major sources of confusion when people talk
and write about anthropomorphism. There seems to be a disconnection between what
people consciously perceive and how they respond to stimuli that they may not
consciously perceive. This means that in the initial phase of interacting with a
novel device \textit{(initialization)} (see Figure \ref{fig:dynamics}) people
might not respond consciously but rather mindlessly \citep{nass_machines_2000},
and in turn anthropomorphize more. They start to respond in a more reflective
manner only during the \textit{familiarization} (in turn, the tendency to
anthropomorphize might decrease). This can be illustrated by the fact that
people tend to deny interacting with computational systems as if they were
people and yet they respond to computers in many ways that are remarkably
similar to how they respond to people \citep{reeves_media_1996}.
\cite{takayama_perspectives_2012} also applies a cognitive viewpoint on the two
different perspectives of in-the-moment \vs reflective to illustrate
differences. She states that in-the-moment perceptions of agency are largely
shaped by bottom-up perceptual processes, evoking very immediate responses (\eg
to the cues emitted by the artifact's design (``artifact-centric
perspective''). In contrast, according to Takayama, reflective perceptions are
more often shaped by top-down processes because of the nature of reflective
thought.

After a longer observation period (typically including complete action sequences
of the robot) or short interaction (touching, short talk like greetings), we
suggest the human activates the cognitive \emph{Layer 2}: in this phase, the human
starts building a behavioral and cognitive model of the robot that would support
both the observed and imagined capabilities of the robot.  The \emph{familiarity
thesis}~\citep{hegel_understanding_2008} supports the idea that the human first
projects onto the robot mental models of similar agents he/she is already
familiar with (ranging from animals to human adults, to pets and children). We
hypothesize that the nature of the projected mental model, as well as how deep
the human engages in this projection, might be driven by the same parameters as
we mentioned for the \emph{initial potential to anthropomorphize}. We also believe
that, at this stage, the human also changes his attitude towards the robot, by
focusing his attention to social cue more than to low-level behaviors, which
may in turn reinforce the resonance mechanisms.

The cognitive \emph{Layer 3} is gradually activated after a
\emph{contextualized} interaction.  A \emph{contextualized} interaction is
\emph{explicitly purposeful} (the purpose of the interaction, be it purely
entertainment, is explicit and conscious to the human), and takes place in an
environment that fosters a stronger cognitive (and possibly affective/social)
commitment from the human in the interaction (typically, at home). During this
interaction, the human iteratively restates and reshapes his/her behavioral and
mental model of the robot (\emph{How does the robot react to such and such
situation/input?  What does the robot know about me? About itself? About our
environment? What can the robot learn?}, etc.).

@@

\draw (-0.3,1.5) node[anchor=east] {Layer 1}
      (-0.3,3) node[anchor=east] {Layer 2}
      (-0.3,4.5) node[anchor=east] {Layer 3};

@@

As mentioned before, the three cognitive stages proposed here do not directly
match with the three phases in the model of the dynamics of anthropomorphism,
and Figure~\ref{fig:mix} attempts to depict their interplay.
The cognitive Layers 1 and 2, in particular, are both active in the
\emph{initialization} phase of the anthropomorphism model. Sub-cognitive
anthropomorphism typically \emph{initiates} the novelty effect by rapidly
engaging the human in the interaction through an initial projected agency,
whereas cognitive Layer 2 (projection of familiar mental models) supports the
novelty effect by inducing beliefs that the robot is set up with possibly
complex cognitive abilities.

The cognitive Layer 3 also overlaps with the \emph{Familiarization} phase: as
the human gets used to the robot, he/she restates and adapts its
cognitive model of the robot by iteratively reshaping pre-existent, familiar
models until it provides a satisfying support to explain and justify the
observed robot behavior.

A \emph{stable level of anthropomorphism} is reached when the adaptation process
depicted in the cognitive Layer 3 reached a stable state, \ie the user's
experience with the robot is correctly supported by the cognitive model he/she
has built.
